---
title: "Agrawal_SoniyaPSET2"
output: pdf_document
date: "2025-09-12"
editor_options: 
  markdown: 
    wrap: sentence
---

Question 1.1 Specify the treatment, observed outcome, and potential outcomes for this causal claim.
Answer: The treatment is immediate access to the program and the control is being put on a wait list.
The treatment effect (Y1-Y0) is the average reduction in shooting victimization due to participation.

Question 1.2 Describe a possible form of selection bias that might affect who ends up participating in the program.
Answer: Given that the program is being offered to people who are "willing to participate", it takes away from the randomness of the treatment effect.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/soniyaagrawal/Desktop/SEM_3/STATS PE/PSET2 data")
my_data <- read.csv("TN_STAR_reduced.csv")
#install.packages("dplyr")
library("dplyr")
#install.packages("ggplot2")
library("ggplot2")
library(knitr)
library(estimatr)


```

```{r pressure, echo=FALSE}
summary(my_data)
head(my_data)
str(my_data)

```

```{r}
#Question 2.1
summary_results <- aggregate(math_score~treat, data = my_data, 
                             FUN = function(x) c(n= length (x), 
                                    mean= mean(x, na.rm =TRUE),
                                   median= median(x, na.rm = TRUE),
                                   sd =sd(x, na.rm = TRUE),
                                   min =min(x, na.rm = TRUE),
                                   max= max(x, na.rm = TRUE)
                           ))
print(summary_results)

#Adding labels
summary_results$Group <- ifelse(summary_results$treat == 0, "Control", "Treatment")

# Reorder columns to put Group first
summary_results <- summary_results[c("Group", "treat", "math_score")]
#view(summary_results)

robust_SE <- lm_robust(math_score~treat, data = my_data)

```

Question 2.2

• What is control group mean?

The control group mean is -0.04933773

• What is the treatment group mean?

The treatement group mean is 0.13046503

• What is the estimated treatment effect?

The treatement effect is Y1-Y0 = 0.1797(0.18)

• What is the estimated standard error on the treatment effect?
What is the “t statistic”?

The SE is 0.03 and the t statistic is 0.18/0.03= 6.

• Is the impact of treat on “math_score” statistically significant at the .05 level?

Our treatment effect is far from 0 is statistically significant.

Question 2.3 Conduct a two-sample t-test for the difference in means between the treatment group and the control group.
What is the t statistic?
Does this differ from your answer to Question 2.2?

Answer: It is the same and the numbers agree.

Question 2.4 Consider the case where we double the sample size.
Holding all else equal, we would see the following:

• Standard Error: Increase, Decrease, or Stay the Same?

The SE will decrease.

• t-statistic: Increase, Decrease, or Stay the Same?

Given that the SE value will shrink, the t stat will increase.

• p-value: Increase, Decrease, or Stay the Same?

With increasing t value, the p-value will become smaller or decrease.

```{r}
#Question 3
Green_data<- read.csv("Gerber_APSR_Reduced.csv")
#View(Green_data)
robust_Green <- lm_robust(vote~treat, data = Green_data)
#View(robust_Green)
# Quick summary
New_data <-aggregate(vote ~ treat, data = Green_data, FUN = function(x) c(n = length(x), mean = mean(x), sd =sd (x), sum = sum(x)))

                
#View(New_data)                    

```

Question 3.1

• What is control group mean?

The control group mean is 0.296

• What is the treatment group mean?

The treatment group mean is 0.3779482.

• What is the estimated treatment effect?

The estimated treatment effect is 0.081.

• What is the estimated standard error on the treatment effect?
What is the “t statistic”?

The estimated SE on the treatment effect is 0.00269.
The t statistic is 30.11

• Is the impact of “treat” on “vote” statistically significant at the .05 level?

There is a significant impact of the treatment on vote.
We have a p value of 4.73e-200.

Question 3.2 Conduct a two-sample t-test for the difference in means between the treatment group and the control group.
What is the t statistic?
Does this differ from your answer to Question 3.1?

Answer: The two supplement each other.

Question 3.3 Consider the case where we double the sample size.
Holding all else equal, we would see the following:

• Standard Error: Increase, Decrease, or Stay the Same?

The SE will decrease by one over the sq root of 2 as our sample size increases.

• t-statistic: Increase, Decrease, or Stay the Same?

The value of t-stat will increase.

• p-value: Increase, Decrease, or Stay the Same?

It will decrease even further.
